### 2017-07-06 paper/compositional.tex
re-structuring sec with compositionality

Compositionality in the context of termination proving means providing the 
proof of termination step-by-step, handling one loop at a time, rather than 
attempting to construct the proof (in our case, a LexRSM) at once. In non-probabilistic setting, the 
relationship between lexicographic ranking functions and compositional 
termination proving is well established, as well as the importance of 
compositional proofs for efficient implementation~\cite{xxx}. In the context of 
probabilistic programs, the work~\cite{HolgerPOPL} attempted to provide a 
compositional notion of almost-sure termination proof based on the 
\emph{probabilistic variant rule (V-rule),} which we explain in a more detail 
below. The reason for this attempt was to provide a method of proving a.s. termination for programs where 1-dimensional ranking supermartingales are not sufficient. Since our LexRSMs were conceived with a similar motivation, we devote this section to a thorough comparison of our approach with the one of~\cite{HolgerPOPL}.

There are two main features of LexRSMs compared with the previous approach, which we call a PVSM (probabilistci V-rule supermartingale) approach. First, the PVSM approach is only applicable to programs in which termination of each loop can be witnessed by a suitable 1-dimensional ranking supermartingale. But there are already single-loop programs for which no 1-dimensional ranking supermartingale exists but a LexRSM does, see the following example.

\textbf{[PETR: ADD EXAMPLE]}

The need for multi-dimensional ranking arises within loops that might go through several ``computation phases'' before terminating, and loops containing non-deterministic choices.

Second, as already noted in~\cite{HolgerPOPL}, the probabilistic V-rule is not 
sound in general. In order to rectify this issue, in~\cite{HolgerPOPL} they 
impose an additional constraint of \emph{uniform integrability} on ranking 
supermartingales that prove termination of individual loops, 
under which the probabilistic V-rule is shown to be sound. Apart from uniform 
integrability being somewhat restrictive in itself, in~\cite{HolgerPOPL} it is 
argued that proving uniform integrability is beyond the capability of 
state-of-the-art automated theorem provers. As a substitute for 
these,~\cite{HolgerPOPL} introduces a type system that can be used to 
automatically prove 
uniform integrability of ranking supermartingales for a restricted class of 
programs. In particular, the method cannot handle programs in which termination 
is controlled by a variable that can be modified by a non-constant value in some sub-program, e.g. 
variable $x$ which appears in an assignment of the form $x:=x-y$. In this 
section we show that using our insights into LexRSMs we 
can obtain a subtly different notion of a probabilistic V-rule which is sound 
without any additional assumptions, and which can be used to compositionally prove termination 
of programs that the previous method cannot handle.

@@

Given an invariant map $\inv$, we say that a 1-dimensional measurable map $\lem$ for $\pCFG_{\program}$ is $\eps$-$\inv$-ranking in location $\loc$, if for each $\vec{x}\in\inv(\loc)$  it holds $\preexp{\lem}(\loc,\vec{x})\leq \lem(\loc,\vec{x})-\eps.$ The map $\lem$ is $\inv$-\emph{unaffected} in location $\loc$ if it is $0$-ranking in $\loc$, and $\inv$-non-negative in $\loc$ if $\lem(\loc,\vec{x})\geq 0$ for all $\vec{x} \in \inv(\loc)$. A measurable map $\lem$ is propositionally linear, if each function $\lem(\loc)$ is of the form $\indicator{G_1}\cdot E_1 + \cdots \indicator{G_k}\cdot E_k$, where each $\indicator{G_i}$ is an indicator function of some polyhedron and each $E_i$ is a linear expression.

We recall the notion of compositional ranking supermartingale as introduced in~\cite{HolgerPOPL}. We call it a PV supermartingale, as it is based on so called probabilistic variant rule. Due to differences in syntax and semantics, the definition is syntactically slightly different from~\cite{HolgerPOPL}, but the essence is the same.

@@

In~\cite{HolgerPOPL} they show that even if all nested loops were already proved to terminate a.s. and there is a PVSM for the program, then the program itself might not terminate a.s. Then they investigate conditions under which existence of a PVSM together with a proof of a.s. termination of each nested sub-loop of $\program$ entails termination of the whole program $\program$. The condition were formulated in terms of a certain type system for \PP{}s, so we say that a PVSM satisfying the condition \emph{typechecks} correctly. In~\cite{HolgerPOPL} the following was proved:

@@

We do not repeat the precise definition of the typesystem here, we just explain its motivation and give examples where it fails, i.e. examples where $\program$ terminates a.s. (and hence the same holds also for its nested loops) and a PV martingale exists, but no PVSM typechecks.

The purpose of the typesystem is, in principle, to ensure that the stochastic 
process which, in each time step, returns the current value of the PVSM, is 
uniformly integrable inside the nested loops. Uniform integrability is a deep 
concept from probability and measure theory: a sequence $X_0,X_1,X_2,\dots$ of 
random variables is uniformly integrable if for each $\delta>0$ there exists an 
$K\in \Nset$ such that for all $n\geq 0$ it holds $\E[|X_n|\cdot\indicator{X_n 
\geq K}]\leq \delta$. As noted in~\cite{HolgerPOPL}, proofs of uniform 
integrability are hard to (semi)-automatize, and hence they devise the 
typechecking as a sound (but not complete) method of proving uniform 
integrability. This is illustrated in the following example.

\begin{example}
Consider the two \APP{}s in Figure~\ref{fig:uniint}, that differ only in one 
coefficient in the 
assignment on line 5. For the inner loop there exists (in both cases) a 
1-dimensional linear ranking supermartingale whose value in each location is 
equal to $c+d_{\loc}$, where $d_{\loc}$ is a location-specific constant. Since 
the expected change of $c$ in each loop step is $-0.5$, this is indeed a LRSM. 
Also, in both cases, a LEM of the form $x+d_{\loc}'$, again for some suitable 
location-specific constants $d_{\loc}'$, is a PVSM for the outer loop, as $x$ 
decreases and is non-negative within the outer loop and its expected change is 
non-negative within the inner loop (more precisely, the inner-loop expected 
change of $x$ zero in the left program and $-\frac{x}{4} - \frac{3}{4}$ in the 
right 
program). 
However, the variable $x$ is uniformly integrable within the inner loop of the 
right program while for the left program this does not hold. To see this, 
imagine the inner loop as a stand-alone program and let $X_n$ be the value of 
variable $x$ after $n$ steps of this stand-alone program (i.e., when the loop 
terminates $x$ no longer changes). Solving a simple linear recurrence shows 
that in the right program $\E[X_n] \rightarrow 0$ as $n\rightarrow \infty$, 
which in particular shows uniform integrability of $X_0,X_1,X_2,\dots$. On the 
other hand, in the left program for each $K>0$ we have $\probm(X_K \geq 
2^K\cdot{x_0})\geq \frac{1}{2^K}$, where $x_0$ is the value of $x$ upon 
entering the loop. Hence $\E[|X_K|cdot\indicator{X_K 
\geq K}]\geq 1$ for each $K$ sufficiently large, which is incompatible with 
uniform integrability. The example shows that proving uniform integrability 
requires intricate reasoning about quantitative behaviour of the program. 
Moreover, as shown below, none of the two programs have a PVSM that typechecks.

@@

\begin{theorem}
\label{thm:nonneg-comp}
Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
\programbody \textbf{ od}$. Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a NCSM for $\program$ supported by some invariant map. Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
\textbf{[PETR: TBD]}
\end{proof}

Hence, NCSMs effectively trade the uniform integrability condition for 
non-negativity over the whole program. We believe that the latter condition is 
substantially easier to impose and check automatically, especially in the case 
of affine probabilistic programs and (propositionally) linear NCSMs: for 
\APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants 
(for which there is a good automated tool support~\cite{xxx}) that 
overapproximate set of reachable configurations and encoding the rankingn, 
unaffection, and non-negativity conditions into a collection of linear 
constraints (as for general LinLexRSMs in Section~\ref{xxx}). 
Figures~\ref{fig:uniint} and~\ref{fig:uniint2} show instances where attempts to 
to prove of a.s. termination via PVSMs fail while proofs via LinNCSMs work.

>>>>

\subsection{One-Dimensional Compositional Proofs of Almost-Sure Termination}

Compositionality in the context of termination proving means providing the 
proof of termination step-by-step, handling one loop at a time, rather than 
attempting to construct the proof (in our case, a LexRSM) at 
once~\cite{KSTW10:compositional-transition-invariants}. 
%In non-probabilistic setting, the 
%relationship between lexicographic ranking functions and compositional 
%termination proving is well established, as well as the importance of 
%compositional proofs for efficient implementation. 
In the context of 
probabilistic programs, the work~\cite{HolgerPOPL} attempted to provide a 
compositional notion of almost-sure termination proof based on the 
\emph{probabilistic variant rule (V-rule),} which we explain in a more detail 
below. However, for the method to work,~\cite{HolgerPOPL} imposes a 
technical~\emph{uniform integrability} condition, whose checking is hard to 
automatize. In this 
section we show that using our insights into LexRSMs we 
can obtain a different notion of a probabilistic V-rule which is sound 
without any additional assumptions, and which can be used to compositionally 
prove termination 
of programs that the previous method cannot handle.

%The reason for this attempt was to provide a method of 
%proving a.s. termination for programs where 1-dimensional ranking 
%supermartingales are not sufficient. Since our LexRSMs were conceived with a 
%similar motivation, we devote this section to a thorough comparison of our 
%approach with the one of~\cite{HolgerPOPL}.

%There are two main features of LexRSMs compared with the previous approach, 
%which we call a PVSM (probabilistic V-rule supermartingale) approach. First, 
%the PVSM approach is only applicable to programs in which termination of each 
%loop can be witnessed by a suitable 1-dimensional ranking supermartingale. But 
%there are already single-loop programs for which no 1-dimensional ranking 
%supermartingale exists but a LexRSM does, see the following example.
%
%\textbf{[PETR: ADD EXAMPLE]}
%
%%The need for multi-dimensional ranking arises within loops that might go 
%%through several ``computation phases'' before terminating, and loops 
%%containing 
%%non-deterministic choices.
%
%Second, as already noted in~\cite{HolgerPOPL}, the probabilistic V-rule is not 
%sound in general. In order to rectify this issue, in~\cite{HolgerPOPL} they 
%impose an additional constraint of \emph{uniform integrability} on ranking 
%supermartingales that prove termination of individual loops, 
%under which the probabilistic V-rule is shown to be sound. Apart from uniform 
%integrability being somewhat restrictive in itself, in~\cite{HolgerPOPL} it is 
%argued that proving uniform integrability is beyond the capability of 
%state-of-the-art automated theorem provers. As a substitute for 
%these,~\cite{HolgerPOPL} introduces a type system that can be used to 
%automatically prove 
%uniform integrability of ranking supermartingales for a restricted class of 
%programs. In particular, the method cannot handle programs in which 
%termination 
%is controlled by a variable that can be modified by a non-constant value in 
%some sub-program, e.g. 
%variable $x$ which appears in an assignment of the form $x:=x-y$.

@@

Given an invariant map $\inv$, we say that a 1-dimensional measurable map 
$\lem$ for $\pCFG_{\program}$ is $\eps$-$\inv$-ranking in location $\loc$, if 
for each $\vec{x}\in\inv(\loc)$  each generalized transition~$\tilde{\tau}$ 
outgoing from $\loc$ is $\eps$-ranked by $\lem$.

We recall the notion of compositional ranking supermartingale as introduced 
in~\cite{HolgerPOPL}. We call it a PV supermartingale, as it is based on so 
called probabilistic variant rule. Due to differences in syntax and semantics, 
the definition is syntactically slightly different from~\cite{HolgerPOPL}, but 
the essence is the same. A measurable map $\lem$ is propositionally linear, if 
each function $\lem(\loc)$ is of the form $\indicator{G_1}\cdot E_1 + \cdots 
\indicator{G_k}\cdot E_k$, where each $\indicator{G_i}$ is an indicator 
function of some polyhedron and each $E_i$ is a linear expression.

@@

In~\cite{HolgerPOPL} they show that even if all nested loops were already 
proved to terminate a.s. and there is a PVSM for the program, then the program 
itself might not terminate a.s. Then they impose a \emph{uniform integrability} 
constraint on the PVSM under which a 
PVSM together with a proof of a.s. termination of each nested sub-loop of 
$\program$ entails termination of the whole program $\program$. 
%they 
%impose an additional constraint of \emph{uniform integrability} on ranking 
%supermartingales that prove termination of individual loops, 
%under which the probabilistic V-rule is shown to be sound. 
Uniform integrability is a deep 
concept from probability and measure theory: a sequence $X_0,X_1,X_2,\dots$ of 
random variables is uniformly integrable if for each $\delta>0$ there exists an 
$K\in \Nset$ such that for all $n\geq 0$ it holds $\E[|X_n|\cdot\indicator{X_n 
	\geq K}]\leq \delta$.
Apart from uniform 
integrability being somewhat restrictive in itself, in~\cite{HolgerPOPL} it is 
argued that proving uniform integrability is beyond the capability of 
state-of-the-art automated theorem provers. As a substitute for 
these,~\cite{HolgerPOPL} introduces a type system that can be used to 
automatically prove 
uniform integrability of ranking supermartingales for a restricted class of 
programs.
We do not repeat the precise definition of the typesystem here, we just say 
that a PVSM satisfying the condition imposed by the type system 
\emph{typechecks} correctly. 
In~\cite{HolgerPOPL} the following was proved:

@@

%, we just explain its motivation and give examples where it fails, i.e. 
%examples where $\program$ terminates a.s. (and hence the same holds also for 
%its nested loops) and a PV martingale exists, but no PVSM typechecks.

%The purpose of the typesystem is, in principle, to ensure that the stochastic 
%process which, in each time step, returns the current value of the PVSM, is 
%uniformly integrable inside the nested loops.  As noted in~\cite{HolgerPOPL}, 
%proofs of uniform 
%integrability are hard to (semi)-automatize, and hence they devise the 
%typechecking as a sound (but not complete) method of proving uniform 
%integrability. This is illustrated in the following example.
The intricacies of uniform integrability are shown in the following example.

\begin{example}
\label{ex:uniform}
\label{EX:UNIFORM}
Consider the two \APP{}s in Figure~\ref{fig:uniint}, that differ only in one 
coefficient in the 
assignment on line 5. For the inner loop there exists (in both cases) a 
1-dimensional linear ranking supermartingale whose value in each location is 
equal to $c+d_{\loc}$, where $d_{\loc}$ is a location-specific constant. Since 
the expected change of $c$ in each loop step is $-0.5$, this is indeed a LRSM. 
Also, in both cases, a LEM of the form $x+d_{\loc}'$, again for some suitable 
location-specific constants $d_{\loc}'$, is a PVSM for the outer loop, as $x$ 
decreases and is non-negative within the outer loop and its expected change is 
non-negative within the inner loop (more precisely, the inner-loop expected 
change of $x$ zero in the left program and $-\frac{x}{4} - \frac{3}{4}$ in the 
right 
program). 
However, the variable $x$ is uniformly integrable within the inner loop of the 
right program while for the left program this does not hold: we show this 
in~\AppendixMaterial. The example shows that proving uniform integrability 
requires intricate reasoning about quantitative behaviour of the program. 
Moreover, as shown below, none of the two programs have a PVSM that typechecks.

@@

\begin{theorem}
\label{thm:nonneg-comp}
Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
\programbody \textbf{ od}$. Assume that each nested loop of $\program$ 
terminates almost surely from each reachable configuration, and that there 
exists a NCSM for $\program$ supported by some invariant map. Then $\program$ 
terminates almost surely from each initial configuration.
\end{theorem}
\begin{proof}[Proof (Key Idea)]
Assume that there is set of non-terminating runs $A$ of positive probability.
Let $\{X_i\}_{i=0}^{\infty}$ be a stochastic process returning the value of 
NCSM $\lem$ in step $i$. We show that for each $i$ it holds $\E[X_i] 
\leq \lem(\locinit,\vecinit) - \eps\cdot \E[\sharp_i]$, where $\sharp_i$ is the 
random variable returning the number of times a location from 
$\slice(\program)$ is visited. Since all sub-loops of $\program$ terminate 
a.s., $\lim_{i\rightarrow \infty}\sharp_i(\run)=\infty$ for $\run\in A$. Since 
$A$ has positive probability $\lim_{i\rightarrow 
\infty}\E[\sharp_i(\run)=\infty]$, it holds $\lim_{i\rightarrow 
\infty}\E[X_i] = -\infty$, a contradiction with non-negativity of $\lem$.
\end{proof}

Hence, NCSMs effectively trade the uniform integrability condition for 
non-negativity over the whole program. We believe that the latter condition is 
substantially easier to impose and check automatically, especially in the case 
of affine probabilistic programs and (propositionally) linear NCSMs: for 
\APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants 
(for which there is a good automated tool support~\cite{FG10:aspic})  encoding 
the ranking, 
unaffection, and non-negativity conditions into a collection of linear 
constraints (as for general LinLexRSMs in Section~\ref{sec:algo}). 
Figures~\ref{fig:uniint} and~\ref{fig:uniint2} show instances where attempts to 
to prove of a.s. termination via PVSMs fail while proofs via LinNCSMs work.

### 2017-07-06 paper/experiments.tex
re-structuring sec with compositionality

\section{Experimental Results}\label{sec:experiments}
We have implemented our algorithm (in C++) and present two sets of experimental results.
For all the experimental results we use the tool Aspic~\cite{} for invariant generation, 
and our algorithm requires linear-programming solver for which we use CPlex~\cite{}.
All our experimental results were obtained on the following platform: 
Ubuntu16.04, 7.7GB, Intel-Core i3-4130 CPU 3.40GHz QuadCore 64-bit.
{\bf KRISH: TWO CITATIONS.}

>>>>

\section{Experimental Results}\label{sec:experiments}
We have implemented our algorithm (in C++) and present two sets of experimental results.
For all the experimental results we use the tool Aspic~\cite{FG10:aspic} for 
invariant generation, 
and our algorithm requires linear-programming solver for which we use 
CPlex~\cite{cplex}.
All our experimental results were obtained on the following platform: 
Ubuntu16.04, 7.7GB, Intel-Core i3-4130 CPU 3.40GHz QuadCore 64-bit.
%{\bf KRISH: TWO CITATIONS.}

### 2017-07-06 paper/introduction.tex
re-structuring sec with compositionality

\noindent{\em Probabilistic programs with nondeterminism.} 
Randomness plays a fundamental role in many areas across science, and in 
computer science in particular.
In applications such as stochastic network protocols~\cite{BaierBook,prism},
randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{LearningSurvey,G15},
the probabilistic behavior must be considered to faithfully model the underlying dynamic system.
The extension of classical imperative programs with \emph{random value generators}, 
that produce random values according to some desired probability distribution, 
naturally gives rise to probabilistic programs.
Along with probability, nondeterminism also plays a crucial role.
In particular in program analysis, for effective analysis of large programs,
all variables cannot be considered, and abstraction ignores some variables,
and the worst-case analysis is represented by adversarial nondeterminism.
Hence, probabilistic programs with nondeterminism have become an active and 
important research focus in program analysis.

>>>>

\noindent{\em Probabilistic programs with nondeterminism.} 
Randomness plays a fundamental role in many areas across science, and in 
computer science in particular.
In applications such as stochastic network protocols~\cite{BaierBook,prism},
randomized algorithms~\cite{RandBook,RandBook2}, 
security~\cite{BGGHS16:diff-privacy-coupling,BGHP16:diff-privacy-siglog} 
machine 
learning~\cite{LearningSurvey,G15},
the probabilistic behavior must be considered to faithfully model the underlying dynamic system.
The extension of classical imperative programs with \emph{random value generators}, 
that produce random values according to some desired probability distribution, 
naturally gives rise to probabilistic programs.
Along with probability, nondeterminism also plays a crucial role.
In particular in program analysis, for effective analysis of large programs,
all variables cannot be considered, and abstraction ignores some variables,
and the worst-case analysis is represented by adversarial nondeterminism.
Hence, probabilistic programs with nondeterminism have become an active and 
important research focus in program analysis.

### 2017-07-06 paper/prelims.tex
re-shuffled the proof of the main theorem a bit
re-structuring sec with compositionality

We use standard notation in the field of probabilistic program 
analysis~\cite{POPL17PAPER}.

@@

\smallskip\noindent{\em Probabilistic Control Flow Graphs.}
%There are several ways in which one can express the semantics of \APP's with
%(angelic and demonic) 
%non-determinism
We consider standard operational semantics of \PP{}s defined via an 
uncountable state-space
Markov decision process (MDP) (uncountable due to real-valued variables).
That is, we associate to each program 
a certain stochastic process.
To define this process, we first define so called 
\emph{probabilistic control flow graphs}~\cite{HONGFEI-CAV16}.

@@

\smallskip\noindent{\em Termination and termination time.}
In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. This 
pCFG has a special location  $\loc_{P}^{\lout}$ corresponding to the value of 
the program counter after 
executing $P$.
We say that a run 
\emph{terminates} if it reaches a 
configuration whose first component is
$\loc_{P}^{\lout}$.
%
We define a random variable $\ttime$
such that for each run $\run$ the value $\ttime(\run)$ represents 
the first 
point in time when 
the current location is $\loc_P^{\lout}$. If a run $\run$ does 
\emph{not} 
terminate, then 
$\ttime(\run)=\infty$. We call $\ttime$ the 
\emph{termination time} of $\program$. 
%In particular, if $\confset$ is the set 
%of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
%terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
%called a \emph{termination time}, as it returns the number of steps after 
%which 
%$\run$ terminates. Since termination time is an important concept on its own, 
%we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

>>>>

We use a standard notation in the field of probabilistic program 
analysis~\cite{CNZ17}.

@@

\smallskip\noindent{\em Probabilistic Control Flow Graphs.}
%There are several ways in which one can express the semantics of \APP's with
%(angelic and demonic) 
%non-determinism
We consider standard operational semantics of \PP{}s defined via an 
uncountable state-space
Markov decision process (MDP) (uncountable due to real-valued variables).
That is, we associate to each program 
a certain stochastic process.
To define this process, we first define so called 
\emph{probabilistic control flow graphs}~\cite{CFG16}.

@@

\smallskip\noindent{\em Termination and termination time.}
In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. This 
pCFG has a special location  $\locterm$ corresponding to the value of 
the program counter after 
executing $P$.
We say that a run 
\emph{terminates} if it reaches a 
configuration whose first component is
$\loc_{P}^{\lout}$.
%
We define a random variable $\ttime$
such that for each run $\run$ the value $\ttime(\run)$ represents 
the first 
point in time when 
the current location is $\loc_P^{\lout}$. If a run $\run$ does 
\emph{not} 
terminate, then 
$\ttime(\run)=\infty$. We call $\ttime$ the 
\emph{termination time} of $\program$. 
%In particular, if $\confset$ is the set 
%of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
%terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
%called a \emph{termination time}, as it returns the number of steps after 
%which 
%$\run$ terminates. Since termination time is an important concept on its own, 
%we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

### 2017-07-06 paper/synthesis.tex
re-structuring sec with compositionality

\section{Algorithmic Aspects}

In this section we describe a polynomial-time algorithm for synthesizing linear $\eps$-LexRSM maps in affine probabilistic programs supported by a given linear invariant map $\inv$. The algorithm, based on iterative solving of linear constraints, is an adaptation of a similar algorithm for finding lexicographic ranking functions in non-probabilistic programs~\cite{ADFG10:lexicographic}. Hence, we provide only a high-level description, focusing on new aspects of our adaptation.

@@

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
follow by arguments identical to those presented 
in~\cite{ADFG10:lexicographic}. 
\begin{theorem}
\label{thm:algo}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. The algorithm runs in time polynomial in size of 
	$\program$ and $\inv$. \textbf{[PETR: POLYHEDRAL GUARDS]}
\end{theorem}
\begin{proof}[Proof (Key Ideas)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$. Non-negativity of each $\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.

Proving completeness is more intricate; as pointed out in~\cite{ADFG10:lexicographic}, one needs to show that the greedy strategy of selecting LEM that ranks maximal number of remaining transitions does not cut off some possible LinLexRSMs. In~\cite{ADFG10:lexicographic} the completeness of the greedy strategy is proved by using several geometric arguments that exploit the fact that the underlying programs are affine. The same geometric properties hold for our generalization to \APP{}s (all ranking conditions in Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program variables), so the result is easily transferable.
\end{proof}

>>>>

\section{Algorithmic Aspects}
\label{sec:algo}

In this section we describe a polynomial-time algorithm for synthesizing linear 
$\eps$-LexRSM maps in affine probabilistic programs supported by a given linear 
invariant map $\inv$. The algorithm, based on iterative solving of linear 
constraints, is a generalization of an algorithm for finding 
lexicographic ranking functions in non-probabilistic 
programs~\cite{ADFG10:lexicographic}. Hence, we provide only a high-level 
description, focusing on the new aspects.

@@

Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
follow by arguments identical to those presented 
in~\cite{ADFG10:lexicographic}. 
\begin{theorem}
\label{thm:algo}
\label{THM:ALGO}
	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
	together with linear invariant map $\inv$. If the algorithm returns a 
	$d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ 
	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
	supported by $\inv$. If guards of all conditionals and loops in $\APP{}$ 
	are linear assertion, then the
	algorithm runs in time polynomial in size of 
	$\program$ and $\inv$.
\end{theorem}
\begin{proof}[Proof (Key Ideas)]
For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$. Non-negativity of each $\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.

Proving completeness is more intricate; as pointed out 
in~\cite{ADFG10:lexicographic}, one needs to show that the greedy strategy of 
selecting LEM that ranks maximal number of remaining transitions does not cut 
off some possible LinLexRSMs. In~\cite{ADFG10:lexicographic} the completeness 
of the greedy strategy is proved by using several geometric arguments that 
exploit the fact that the underlying programs are affine. The same geometric 
properties hold for our generalization to \APP{}s (all ranking conditions in 
Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program 
variables), so the result is easily transferable. The complexity argument is 
rather standard and we present it in~\AppendixMaterial.
\end{proof}
